Omics-based Single Step Trait Prediction

### 2017-09-14 paper/manuscript.tex
paper, fix typos

\Genetics2level{Exp2}
To address objectives 1 and 2, a maize hybrid data set, subsequently
denoted as Exp2, was used.
The material comprised 1,521 hybrids produced in 16 factorial mating designs
between 142 Dent and 103 Flint parent lines (described in details by
\citeNP{Westhues2017}).
Best linear unbiased estimates (BLUEs) were computed across all factorials and 
three or more agro-ecologically diverse environments across Germany for seven
agronomically important traits in silage maize by accounting for year, location,
field replication, block and genotype effects as well as their interactions
(Table~\ref{table:TraitKey}).
All 245 parent lines have pedigree records back to the generation of their 
grandparents or deeper~\cite{Westhues2017} and were genotyped using the Illumina 
BeadChip MaizeSNP50~\cite{Ganal2011}.
The same SNP quality checks as for the material in Exp1 were applied here,
yielding 7,013 polymorphic marker loci for the Dent and 6,212 for the Flint
lines, respectively.
Gene expression data for seedlings seven days after sowing were obtained for a 
subset of 60 Dent and 43 Flint lines, that are parents of 685 hybrid progeny.
These data were obtained through two-color hybridizations using a custom 2K 
microarray (GPL22267)~\cite{Westhues2017}.
After applying established normalization procedures~\cite{Smyth2003,Ritchie2007},
BLUEs for gene expression were computed for 1,323 transcripts
\cite{Westhues2017} using the R-package \emph{limma}~\cite{Ritchie2015a}.

@@

Consider now the situation in which one predictor is complete in the sense that
the whole population is equipped with features for that predictor whereas only
a subset of that population has information on a second predictor.
We can take a similar route as in \citeNP{Fernando2014} and impute covariates 
of the incomplete predictor by using covariates of the complete predictor.
Let the subscript $1$ denote individuals covered only by the complete
predictor whereas individuals covered by both, the complete and the
incomplete predictor, are indicated by the subscript $2$.
Let us further denote the matrix of centered covariates of the incomplete
predictor as
\begin{equation} \label{eq:incomplete-covariates}
\mathbf{W} =
\begin{bmatrix}
  \mathbf{Z}_{1}\mathbf{\hat{W}}_{1} \\
  \mathbf{W}_{2}
 \end{bmatrix},
\end{equation}
where observed feature covariates are stored in $\mathbf{W}_{2}$ while marker
covariates for genotypes not covered by the incomplete predictor
($\mathbf{\hat{W}}_{1} = \mathbf{K}_{12} \mathbf{K}^{-1}_{22} \mathbf{W}_{2}$)
need to be imputed.
In this case we need to partition the vector of genetic values $\mathbf{g}$
into a component $\mathbf{g}_{1}$ (\textit{i.e.,} genetic values of genotypes
covered only by the complete predictor) and a component
$\mathbf{g}_{2} = \mathbf{W}_{2}\boldsymbol{\alpha}$ (\textit{i.e.,} genetic
values of genotypes covered by both predictors where $\boldsymbol{\alpha}$
pertains to the random feature effects of the incomplete predictor).
Under multivariate normality, the vector $\mathbf{g}_1$ can be written as the
sum of the conditional expectation given $\mathbf{g}_2$ and a residual error
term:
\begin{align} \label{eq:mrna1}
	\mathbf{g_1} &= E(\mathbf{g}_1|\mathbf{g}_2) + \boldsymbol{\epsilon} \\
	&= \mathbf{K}_{12}\mathbf{K}^{-1}_{22}\mathbf{W}_{2}
	\boldsymbol{\hat{\alpha}} + (\mathbf{g_1} - \mathbf{K}_{12}
	\mathbf{K}^{-1}_{2}
	\boldsymbol{\hat{\alpha}}) \\
	&= \mathbf{\hat{g}}_1 + \boldsymbol{\epsilon},
\end{align}

@@

Extending this model to hybrids with parent lines from two heterotic groups
simply requires us to separate the feature matrices of the two predictors
into a paternal and a maternal component, respectively.
Using the index $D$ for the Dent and the index $F$ for the Flint lines in Exp2
we then have a feature matrix $\mathbf{M}_{D}$ and $\mathbf{M}_{F}$ for each
predictor.
For the hybrids the final model is therefore simply
\begin{align} \label{eq:final-model}
\mathbf{y} &= \mathbf{Xb} +
\mathbf{Z}_{D}\mathbf{W}_{D} \boldsymbol{\alpha}_{D} + 
\mathbf{Z}_{D}\mathbf{U}_{D} \boldsymbol{\epsilon}_{D} +
\mathbf{Z}_{F}\mathbf{W}_{F} \boldsymbol{\alpha}_{F} + 
\mathbf{Z}_{F}\mathbf{U}_{F} \boldsymbol{\epsilon}_{F} +
\mathbf{e}.
\end{align}

>>>>

\Genetics2level{Exp2}
To address objectives 1 and 2, a maize hybrid data set, subsequently
denoted as Exp2, was used.
The material comprised 1,521 hybrids produced in 16 factorial mating designs
between 142 Dent and 103 Flint parent lines (described in detail by
\citeNP{Westhues2017}).
Best linear unbiased estimates (BLUEs) were computed across all factorials and 
three or more agro-ecologically diverse environments across Germany for seven
agronomically important traits in silage maize by accounting for year, location,
field replication, block and genotype effects as well as their interactions
(Table~\ref{table:TraitKey}).
All 245 parent lines have pedigree records back to the generation of their 
grandparents or deeper~\cite{Westhues2017} and were genotyped using the Illumina 
BeadChip MaizeSNP50~\cite{Ganal2011}.
The same SNP quality checks as for the material in Exp1 were applied here,
yielding 7,013 polymorphic marker loci for the Dent and 6,212 for the Flint
lines, respectively.
Gene expression data for seedlings seven days after sowing were obtained for a 
subset of 60 Dent and 43 Flint lines, that are parents of 685 hybrid progeny.
These data were obtained through two-color hybridizations using a custom 2K 
microarray (GPL22267)~\cite{Westhues2017}.
After applying established normalization procedures~\cite{Smyth2003,Ritchie2007},
BLUEs for gene expression were computed for 1,323 transcripts
\cite{Westhues2017} using the R-package \emph{limma}~\cite{Ritchie2015a}.

@@

Consider now the situation in which one predictor is complete in the sense that
the whole population is equipped with features for that predictor whereas only
a subset of that population has information on a second predictor.
We can take a similar route as in \citeNP{Fernando2014} and impute covariates 
of the incomplete predictor by using covariates of the complete predictor.
Let the subscript $1$ denote individuals covered only by the complete
predictor whereas individuals covered by both, the complete and the
incomplete predictor, are indicated by the subscript $2$.
Let us further denote the matrix of centered covariates of the incomplete
predictor as
\begin{equation} \label{eq:incomplete-covariates}
\mathbf{W} =
\begin{bmatrix}
  \mathbf{Z}_{1}\mathbf{\hat{W}}_{1} \\
  \mathbf{W}_{2}
 \end{bmatrix},
\end{equation}
where observed feature covariates are stored in $\mathbf{W}_{2}$ while feature
covariates for genotypes not covered by the incomplete predictor
($\mathbf{\hat{W}}_{1} = \mathbf{K}_{12} \mathbf{K}^{-1}_{22} \mathbf{W}_{2}$)
need to be imputed.
In this case we need to partition the vector of genetic values $\mathbf{g}$
into a component $\mathbf{g}_{1}$ (\textit{i.e.,} genetic values of genotypes
covered only by the complete predictor) and a component
$\mathbf{g}_{2} = \mathbf{W}_{2}\boldsymbol{\alpha}$ (\textit{i.e.,} genetic
values of genotypes covered by both predictors where $\boldsymbol{\alpha}$
pertains to the random feature effects of the incomplete predictor).
Under multivariate normality, the vector $\mathbf{g}_1$ can be written as the
sum of the conditional expectation given $\mathbf{g}_2$ and a residual error
term:
\begin{align} \label{eq:mrna1}
	\mathbf{g_1} &= E(\mathbf{g}_1|\mathbf{g}_2) + \boldsymbol{\epsilon} \\
	&= \mathbf{K}_{12}\mathbf{K}^{-1}_{22}\mathbf{W}_{2}
	\boldsymbol{\hat{\alpha}} + (\mathbf{g_1} - \mathbf{K}_{12}
	\mathbf{K}^{-1}_{2}
	\boldsymbol{\hat{\alpha}}) \\
	&= \mathbf{\hat{g}}_1 + \boldsymbol{\epsilon},
\end{align}

@@

Extending this model to hybrids with parent lines from two heterotic groups
simply requires separating the feature matrices of the two predictors into a
paternal and a maternal component, respectively.
Using the index $D$ for the Dent and the index $F$ for the Flint lines in Exp2
we then have a feature matrix $\mathbf{M}_{D}$ and $\mathbf{M}_{F}$ for each
predictor.
For the hybrids the final model is therefore simply
\begin{align} \label{eq:final-model}
\mathbf{y} &= \mathbf{Xb} +
\mathbf{Z}_{D}\mathbf{W}_{D} \boldsymbol{\alpha}_{D} + 
\mathbf{Z}_{D}\mathbf{U}_{D} \boldsymbol{\epsilon}_{D} +
\mathbf{Z}_{F}\mathbf{W}_{F} \boldsymbol{\alpha}_{F} + 
\mathbf{Z}_{F}\mathbf{U}_{F} \boldsymbol{\epsilon}_{F} +
\mathbf{e}.
\end{align}
